Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-02 doc/papers/cgo2014/paper.tex
Some slight improvements

PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
enhance Python performance. However Shed~Skin does not provide support for the
Numpy module. The is an important work in progress from the PyPy community to
support Numpy, but it implies a partial rewrite of the Numpy API in RPython,
the language used by PyPy, to enable PyPy's optimization.

@@

Many C++ linear algebra libraries~\cite{eigen,ublas} use \emph{expression
templates}~\cite{expression_templates, et2012} as a way to efficiently describe high
level matrix expressions while avoid the creation of intermediate storage.
Additionally, it has the same effect as loop fusion, solving the issues
described in \S~\ref{sec:temporaries-elimination} and
\S~\ref{sec:operator-fusion}.

@@

\subsection{Range Analysis}

@@

A first exploration of vectorization of Python constructs is done in~\cite{wpmvp2014}.

@@

\subsection{The Pythran Compiler}
\cite{pythran2013,isocxx11}

@@

    \centering
    \begin{tikzpicture}[
            file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
            node distance=0.7cm},
            tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
        distance=0.7cm}]
        \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
        \node[tool] (pythran) [below=of python] {\textbf{Pythran}};
        \node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
        \node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{g++}};
        \node (empty) [xshift=-1em, left=of gxx] {};
        \node[file] (pythonic) [left=of empty] {\textbf{\texttt{pythonic}}};
        \node[file] (annotation)     [above=of pythonic] {\textbf{Type Info}};
        \node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
        \node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module [\texttt{.so}]}};

>>>>

PyPy~\cite{pypy2009}, a fully compatible Python interpreter with a tracing JIT,
or Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler, are also viable
ways to enhance Python performance. However Shed~Skin does not provide support
for the Numpy module. PyPy faces the issue of calling native libraries that
cannot be optimized by a Python JIT compiler. There is an important work in
progress from the community to support Numpy, but it implies a partial rewrite
of the Numpy API in RPython, the language used by PyPy.

@@

Many C++ linear algebra libraries~\cite{eigen,ublas,esterie2014} use \emph{expression
templates}~\cite{expression_templates, et2012} as a way to efficiently describe high
level matrix expressions while avoid the creation of intermediate storage.
Additionally, it has the same effect as loop fusion, solving the issues
described in \S~\ref{sec:temporaries-elimination} and
\S~\ref{sec:operator-fusion}.

@@

%\subsection{Range Analysis}

@@

A first exploration of vectorization of Python constructs is done
in~\cite{wpmvp2014}. It shows that language constructs like list comprehension
offer many vectorization opportunities. This section focuses on the
vectorization of complex Numpy expressions. In order to vectorize a numpy
expression, the last dimension of the expression must be contiguous. This can
be checked at compile time using type traits. The general form of a subscripted
array can be expressed in C++ using a template type, say \texttt{numpy\_gexpr}.
For instance the Numpy subscript \lstinline|a[:,0,::2]| on a 3D float matrix is
represented by the C++ type in Listing~\ref{lst:gexpr}.

\begin{lstlisting}[language=c++, caption={C++ type corresponding to the Numpy subscript \lstinline|a[:,0,::2]| on a 3D float matrix}, label={lst:gexpr}]
numpy_gexpr<ndarray<double, 3>,
            contiguous_slice,
            long, slice>
\end{lstlisting}

This representation makes it easy to
statically compute, in a vectorizable type trait, wether the last dimension of
a subscript is contiguous or not. It can be generalized to the other array
manipulations. BOOST.simd~\cite{esterie2012boost} provides a
machine-independant abstraction for vector operation and supports all the C99
math library. Thanks to this abstraction, all Numpy binary and unary functions
and operators can be vectorized.

The loop described in Listing~\ref{lst:eval-et} can be specialized using this
vectorizable type trait, to automatically vectorize the innermost loop whenever
it is legal. The choice is statically made at compile time.

@@

\subsection{The Pythran Compiler}

The ideas described in this paper are implemnted in the Pythran
compiler~\cite{pythran2013}, a translator from a subset of Python to
C++11~\cite{isocxx11}. The architecture of the compiler is briefely described
in Figure~\ref{fig:pythran-compiler}. The input is a Python module written in
the Pythran subset. Pythran translates it in its internal representation, a
simplified Python AST. It performs various optimization then outputs either
Python code, in a source-to-source fashion, or C++ template code that makes
call to the pythonic library that typically implements the \texttt{ndarray}
interface. User annotations can be used to instanciate this code for the proper
types and generate a native library. This library relies on Boost.Python
library to match Python's C API.

The Pythran compiler is an open source project publicly released under the BSD
license~\footnote{\url{http://pythonhosted.org/pythran/}}.

@@

    \centering
    \begin{tikzpicture}[
            file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
            node distance=0.7cm},
            tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
        distance=0.7cm}]
        \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
        \node[tool] (pythran) [below=of python] {\textbf{Pythran}};
        \node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
        \node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{c++}};
        \node (empty) [xshift=-1em, left=of gxx] {};
        \node[file] (pythonic) [left=of empty] {\textbf{\texttt{pythonic}}};
        \node[file] (annotation)     [above=of pythonic] {\textbf{Type Info}};
        \node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
        \node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module [\texttt{.so}]}};
